---
title: "Hotel Bookings EDA"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file=file.path(dirname(input_file),
    out_dir, 'index.html'))})
author: "Sergio Puno"
date: "2/23/2022"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    theme: cosmo
    collapsed: true
    df_print: paged
editor_options: 
  chunk_output_type: console
---

Source:
https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-02-11/readme.md

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Preparations {.tabset}
Below are the required packages used if you'd like to reproduce the results. The full
source code is available at [my github repository](https://github.com/Sergio-Puno).

## Load libraries
Libraries for data wrangling, general visualization and extra specialized tools.

```{r}
# data manipulation
library(tidyverse)
library(tidytuesdayR)
library(lubridate)
library(tidytext)
library(rlang)

# visualization
library(ggplot2)
theme_set(theme_minimal())
library(scales)

# forecast

# formatting
library(kableExtra)

```

## Helper functions
First create functions that can help with some plotting to reduce repeated code chunks.

```{r}
# quick histogram plotting with median feed-in
plot_distribution <- function(df, feature, x_intercept) {
  x <- ggplot(data = df, aes(x = {{feature}})) +
    geom_histogram(fill = "steelblue3") +
    geom_vline(xintercept = as.double(x_intercept), color = "red")
  
  return(x)
}

# quick bar plot for univariate analysis
plot_bar <- function(df, feature, agg_n) {
  x <- ggplot(data = df, aes(x = {{agg_n}}, y = {{feature}}, fill = {{feature}})) +
    geom_col() +
    theme(legend.position="none")
  
  return(x)
}
```

## Load data
Import file from TidyTuesday repo and local country abbreviations cross reference csv.

```{r}
hotels <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-11/hotels.csv')

country_codes <- read_csv("country_codes.csv")

head(hotels)
```

# Data Background
This information comes from two datasets of hotel demand data, one a resort hotel
in the region of Algarve and the other a city hotel in the region of Lisbon (both in Portugal).

The brief associated with this data can be found [here](https://www.sciencedirect.com/science/article/pii/S2352340918315191#f0010).
One consideration mentioned in this briefing was the below:
<br />

*"...even though the capture of data took considered a timespan prior to arrival date, it is understandable that the distribution of some variables differ between non canceled and canceled bookings."*
<br />

We'll have to consider the impact of a cancelled booking on each distribution and how
a cancelled booking is different (on average) compared to a booking that showed up.

# Data Cleaning / Prelimary Analysis
## Missing Values
Quick missing values check to see if there are any massive holes in the data:

```{r}
# checking for missing values
missing_values <- hotels %>% 
  summarise(across(everything(), ~sum(is.na(.))))

missing_values <- gather(missing_values, key = "feature", value = "missing_pct")

missing_values %>%
  ggplot(aes(x = reorder(feature, missing_pct), y = missing_pct)) +
  geom_bar(stat = "identity", fill = "tomato2") +
  coord_flip()
```

<br />

Looks clean, only the children records seem to have missing values, about 4% of the records
are missing child data, this is not a huge deal so we can keep these records.

### Creating/Updating data fields
Generating any new fields/columns that I feel may be helpful in subsequent analysis
as well as changing some data types from the initial import such as converting certain
fields to factors.

```{r}
# creating a new date column with the year/month/day of month cols for arrivals
hotels <- hotels %>% 
  mutate(arrival_date = dmy(paste(arrival_date_day_of_month,
                                  arrival_date_month,
                                  arrival_date_year,
                                  sep = "-"))) %>% 
  mutate(is_canceled = as.factor(is_canceled),
         stays_in_weekend_nights = as.factor(stays_in_weekend_nights),
         stays_in_week_nights = as.factor(stays_in_week_nights),
         meal = as.factor(meal),
         market_segment = as.factor(market_segment),
         distribution_channel = as.factor(distribution_channel),
         reserved_room_type = as.factor(reserved_room_type),
         assigned_room_type = as.factor(assigned_room_type),
         deposit_type = as.factor(deposit_type),
         customer_type = as.factor(customer_type),
         reservation_status = as.factor(reservation_status))
```

## Univariate Analysis {.tabset}
Distributions and getting to know our data.

<br />

A few things I have in mind while reviewing the distributions, many of these apply
to all distributions I'll look at and not just lead-time specific:
<ul>
<li> How extreme are outliers compared to the majority of the data's distribution
<li> Check for any bimodal distributions or any cycles
<li> Would the cancellation vs. non-cancellation factor impact the distributions
<li> General assessment of the feature's value as a predictor of either the average
daily rate, or for predicting a cancellation
</ul>

You can perform univariate analysis and simple EDA on each feature in order to gain some insight into, for example, categorical features that display low entropy or class
imbalance. These all will inform your model so the features should be reviews for 
skewness and normality.

### Market Segments
Checking most common segments that drive bookings, which appears to be online travel
agents by a large margin with offline travel agent/tour operator and groups following.
Aviation and complementary markets are extremely low (less than 1% of total bookings
combined).

```{r}
hotels %>% 
  filter(market_segment != "Undefined") %>% 
  count(market_segment, sort= TRUE) %>% 
  plot_bar(., market_segment, n) +
  labs(x = "# of Bookings", y = "Market Segment",
       title = "Number of Bookings by Market Segment") +
  scale_x_continuous(labels = comma)

hotels %>% 
  group_by(market_segment) %>% 
  summarise(count = n()) %>% 
  mutate(freq = round(count / sum(count), 3)) %>% 
  arrange(desc(freq))
```

### Countries of visitors
Identifying most common customer country of origin, bringing in full country name from
the helper file I loaded in for country abbreviations. Trips within Portugal are
the most common at 40% of total bookings, and the UK/France/Spain/Germany bringing
up the top 5.

```{r}
hotels %>% 
  left_join(., country_codes, by = "country") %>% 
  filter(!is.na(country_name)) %>% 
  mutate(country_name = fct_lump(country_name, 20)) %>% 
  count(country_name, sort= TRUE) %>% 
  mutate(country_name = fct_reorder(country_name, n)) %>% 
  filter(country_name != "Other") %>% 
  plot_bar(., country_name, n) +
  labs(x = "# of Bookings", y = "Country of Origin",
       title = "Number of Bookings by Country") +
  scale_x_continuous(labels = comma)

hotels %>% 
  group_by(country) %>% 
  summarise(count = n()) %>% 
  mutate(freq = round(count / sum(count), 3)) %>% 
  arrange(desc(freq)) %>% 
  head(10)
```

### Customer Type
What is the most common customer type, we can study this in combination with other
details we have such as `reserved_room_type` or `deposit_Type` to see if certain
types of customers are aligned with certain other requirements.

```{r}
hotels %>% 
  count(customer_type, sort = TRUE) %>% 
  plot_bar(., customer_type, n) +
  labs(x = "Customer Type", y = "",
       title = "Customer Types")

hotels %>% 
  group_by(customer_type) %>% 
  summarise(count = n()) %>% 
  mutate(freq = round(count / sum(count), 3)) %>% 
  arrange(desc(freq))
```

### ADR
Average Daily Rate - this is the only value field we have related to cost in this 
dataset, with this we can take a look later on at how where some of the rate is 
derived from (such as the room type, market segment, meal type, etc.)

```{r}
adr_median <- hotels %>% 
  filter(adr > 0) %>% 
  summarise(median = median(adr))

hotels %>% 
  filter(adr > 0) %>% 
  plot_distribution(., adr, adr_median) +
  scale_x_log10(labels = dollar) +
  labs(x = "ADR (median in red)",
       y = "Occurences",
       title = "Distribution of ADR values above $0")

# top 10 ADR values
hotels %>% 
  arrange(desc(adr)) %>% 
  head(10) %>% 
  select(hotel, is_canceled, arrival_date, assigned_room_type, adr)
```

### Lead Time
Taking a look at the `lead_time` and arrival days to breakdown the bookings , we can split the lead times into two 
categories, same day bookings and standard ahead of date bookings.

```{r}
# median for non 0 LT (lead time) bookings
nonzero_LT_median <- hotels %>% 
  filter(lead_time > 0) %>% 
  summarise(median = median(lead_time))

# for non same-day bookings, was does the lead time distribution look like
hotels %>% 
  filter(lead_time > 0) %>% 
  plot_distribution(., lead_time, nonzero_LT_median) +
  labs(x = "Booking Lead Time", y = "Occurences",
       title = "Lead Time Distribution (nonzero leadtimes") 
```

### Assigned Room

```{r}
hotels %>% 
  group_by(hotel, assigned_room_type) %>%
  summarise(count = n()) %>% 
  mutate(assigned_room_type = reorder_within(assigned_room_type, count, hotel)) %>% 
  ggplot(aes(count, assigned_room_type, fill = assigned_room_type)) +
  geom_col() +
  theme(legend.position = "none") +
  labs(x = "Occurences",
       y = "Assigned Room Type",
       title = "Frequency of Assigned Room Types") +
  facet_wrap(~ hotel, scales = "free_y") +
  scale_y_reordered()
```

<br />

# Market Segment Analysis and Impact

### Market Segment Breakdown
Checking out the top market segments of each hotel.

```{r}
# data manipulation for plotting
hotels_markets <- hotels %>% 
  filter(market_segment != "Undefined") %>% 
  mutate(market_segment = fct_lump(market_segment, 4)) %>%
  count(hotel, market_segment, is_canceled) %>% 
  group_by(hotel) %>% 
  mutate(total = sum(n),
         ratio = n / total,
         is_canceled = ifelse(is_canceled == 0, "No", "Yes")) %>% 
  ungroup() %>% 
  mutate(market_segment = reorder_within(market_segment, ratio, hotel))

hotels_markets %>% 
  ggplot(aes(ratio, market_segment, fill = is_canceled)) +
  geom_col() +
  labs(x = "Percentage of Bookings", y = "Market Segment",
       title = "Hotel Bookings by Market Segment: Hotel Split",
       fill = "Booking Canceled") +
  facet_wrap(~ hotel, scales = "free_y") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = c("steelblue2", "tomato2")) +
  scale_y_reordered() +
  theme(legend.position = "top")
```

<br />

Looking at the results, we can see:
<ul>
<li> The majority of people use travel agents (both online and offline) though we don't know the breakdown of offline agents to tour operators. 
<li> `Direct` is a close follow up for the resort hotel whereas `Groups` is the follow up for the city hotel. 
<li> Cancellations are quite proportionally high for the city hotel segments, especially for the group bookings which makes sense as there is a high chance for one individual out of a group to cause a cancellation due to external factors.
</ul>

Let's take a look at these markets for these two hotels in combination to get a sense
of the overall bookings pool and cancellations.

```{r}
hotels_markets <- hotels %>% 
  filter(market_segment != "Undefined") %>% 
  group_by(market_segment) %>% 
  count(market_segment, is_canceled) %>%
  mutate(total = sum(n),
         ratio = n / total,
         is_canceled = ifelse(is_canceled == 0, "No", "Yes")) %>% 
  arrange(desc(n))

# stacked column chart grouping all bookings
hotels_markets %>% 
  ggplot(aes(market_segment, ratio, fill = is_canceled)) +
  geom_col() +
  labs(x = "Market Segment", y = "Percentage of Bookings",
     title = "Hotel Bookings by Market Segment: Combined",
     fill = "Booking Canceled") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_x_reordered() +
  scale_fill_manual(values = c("steelblue2", "tomato2")) +
  theme(legend.position = "top") +
  coord_flip()
```

<br />

We can see from this stacked bar chart that overall, for bookings using online 
travel agents there are between 15% to almost 20% cancellations. Interestingly, 
compared to the online agents, the `Direct` and `Complementary` bookings have the
lowest cancellation rates with `Corporate` following closely behind. 

<br />


### Arrival date analysis:
Looking at the breakdown of arrival day-of-week by market segment first.

```{r}
# looking at the different market segments and arrival day of week for comparisons

hotels %>% 
  mutate(wday = wday(arrival_date, label = TRUE, week_start = 1)) %>%
  filter(market_segment != "Undefined") %>% 
  group_by(wday, market_segment) %>% 
  count(wday) %>% 
  ggplot(aes(market_segment, y= n, color= wday)) +
  geom_point(size=4) +
  theme(legend.position = "left",
        plot.title = element_text(size=14)) +
  coord_flip() +
  scale_x_discrete(position="to") +
  labs(x = "") +
  ggtitle("Arrival by Weekday") +
  scale_color_hue() +
  scale_y_log10()
```

<br />

Looking at the results we find:
<ul>
<li> `Online TA`, `Offline TA/TO`, `Groups` appear to be the top three markets;
TA = Travel Agent, TO = Tour Operators
<li> You can see that the most common day of week shifts from segment to segment
with Friday being popular for `Online TA` and `Offline TA/TO` but Monday-Wednesday
being more common for `Corporate` which makes sense as if you are working you'll
likely be staying during just the work week
</ul>

<br />

### Breakdown by day of month
We can see that the first week fo the month is universally lower in each market
when compared to say the end of the month which is the top count for all markets 
except for `Aviation` which has the third week as their peak.
```{r}
hotels %>% 
  mutate(month_week = case_when(arrival_date_day_of_month >= 21 ~ '4th WK',
                                arrival_date_day_of_month >= 14 ~ '3rd WK',
                                arrival_date_day_of_month >= 7 ~ '2nd WK',
                                arrival_date_day_of_month < 7 ~ '1st WK')) %>% 
  filter(market_segment != "Undefined") %>%
  group_by(month_week, market_segment) %>% 
  count(month_week) %>% 
  ggplot(aes(market_segment, y= n, color= month_week)) +
  geom_point(size=4) +
  theme(legend.position = "left",
        plot.title = element_text(size=14)) +
  coord_flip() +
  scale_x_discrete(position="to") +
  labs(x = "") +
  ggtitle("Arrival by Week of Month") +
  scale_color_hue() +
  scale_y_log10()
```

<br />

Nothing too spread out for this plot, the main consistency is that the last week
of the month is almost always the number one week, with `Aviation` having the
second week as their most common. We can change this up a bit to see if any months
in particular show trends of bookings.

# Bookings Time Series Analysis
### Time Series of Bookings
First we'll take a look at a single year available in the dataset which spans from
July 1st, 2015 to August 31st, 2017. Looking at 2016 for both the city and resort 
hotels, we can make a few observations:

```{r}
hotels %>% 
  filter(between(arrival_date, as.Date("2016-01-01"), as.Date("2016-12-31"))) %>% 
  group_by(hotel, arrival_date, arrival_date_year) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(arrival_date, count)) +
  geom_line(aes(color = hotel), size = 1) +
  facet_wrap(~ hotel, ncol = 1) +
  theme(legend.position = "none") +
    labs(x = "Arrival Date", y = "# of Bookings",
       title = "Number of Bookings (2016)")
```

<br />
<ul>
<li> The city hotel bookings are more varied as there are more total bookings
(almost double) compared to the resort hotel, but we can see bookings slowly rising
during the first quarter of the year with a spike between Feb and Mar.
<li> If you look at the resort bookings, we see the same spike between Feb and Mar, 
referencing holidays that take place we can find Carnival (Mardi Gras) taking place 
during this time window.
time window a
<li> While I would have expected bookings at a resort to jump in the summer, it appears
that the period of time with the most spikes in bookings are from Feb to May which is
more of a spring time focus.
<li> The final note is the expected spike for new years in both, but this spike for
city hotels is not a massive spike comparatively while the spike for resorts is fairly
large compared to the rest of the year.
</ul>

<br />

```{r}
hotels %>% 
  group_by(hotel, arrival_date_year, arrival_date_week_number) %>% 
  count() %>% 
  group_by(hotel, arrival_date_week_number) %>% 
  summarise(avg_bookings = mean(n)) %>% 
  mutate(year_month = ymd("2000-01-01") + weeks(arrival_date_week_number - 1)) %>% 
  ggplot(aes(year_month, avg_bookings)) +
  geom_line(aes(color = hotel), size = 2) +
  scale_x_date(date_labels = "%b-%d") +
  facet_wrap(~ hotel, ncol = 1) +
  theme(legend.position = "none") +
  labs(x = "Arrival Date", y = "Avg Bookings",
       title = "Average bookings",
       subtitle = "July 2015 - August 2017")
  
```

<br />
Overall this average over the time periods provided shows a bit more clearly the 
trends that each hotel has and how similar/different they can be. Both have a steady 
climb at the beginning of the year with Carnival bookings leading into a dip from
July to September and a final increase at the end of the year for new years.

There is some divergence with the city hotel in the second half of the year 
having a large resurgence in bookings from the end of September to November which 
we'll have to investigate further to identify what customers are driving this increase.

### Time Series Bookings vs. Market Segment
Let's try to break down the avg bookings as we did before but break it down by the 
market segment to see the seasonality that drives business by customer type.

```{r fig.width=15, fig.height=10}
# setting up colors to use for each plot with same market to color
cbbPalette <- c("Online TA" = "#000000", "Offline TA/TO" = "#E69F00", 
                "Groups" = "#56B4E9", "Direct" = "#009E73", "Corporate" = "#F0E442",
                "Complementary" = "#0072B2", "Aviation" = "#D55E00")

hotels %>% 
  filter(hotel == "City Hotel") %>% 
  filter(market_segment != "Undefined") %>% 
  group_by(market_segment, arrival_date_year, arrival_date_week_number) %>% 
  count() %>% 
  group_by(market_segment, arrival_date_week_number) %>% 
  summarise(avg_bookings = mean(n)) %>% 
  mutate(year_month = ymd("2000-01-01") + weeks(arrival_date_week_number - 1)) %>%
  arrange(market_segment) %>% 
  ggplot(aes(year_month, avg_bookings)) +
  geom_line(aes(color = market_segment), size = 1) +
  scale_x_date(date_labels = "%b-%d") +
  theme(legend.position = "bottom",
        legend.justification = "left",
        legend.direction = "horizontal",
        legend.key.size = unit(1.5, "cm")) +
  labs(x = "Arrival Date", y = "Avg Bookings",
       title = "Average bookings: City Hotel vs. Market Segment",
       subtitle = "July 2015 - August 2017",
       color = "") +
  guides(color = guide_legend(nrow = 1)) +
  scale_color_manual(values = cbbPalette)
```

<br />
Based on this breakdown we can see that for the majority of the start of the year
the bookings for the city hotel are driven largely by online travel agents with a 
large spike in bookings through offline travel agents or tour operators just after
October.

Overall the group of Direct, Corporate, Complementary, and Aviation don't contribute
as much as the other categories. 

```{r fig.width=15, fig.height=10}
cbbPalette <- c("Online TA" = "#000000", "Offline TA/TO" = "#E69F00", 
                "Groups" = "#56B4E9", "Direct" = "#009E73", "Corporate" = "#F0E442",
                "Complementary" = "#0072B2")

hotels %>% 
  filter(hotel == "Resort Hotel") %>% 
  filter(market_segment != "Undefined") %>% 
  group_by(market_segment, arrival_date_year, arrival_date_week_number) %>% 
  count() %>% 
  group_by(market_segment, arrival_date_week_number) %>% 
  summarise(avg_bookings = mean(n)) %>% 
  mutate(year_month = ymd("2000-01-01") + weeks(arrival_date_week_number - 1)) %>%
  arrange(market_segment) %>% 
  ggplot(aes(year_month, avg_bookings)) +
  geom_line(aes(color = market_segment), size = 1) +
  scale_x_date(date_labels = "%b-%d") +
  theme(legend.position = "bottom",
        legend.justification = "left",
        legend.direction = "horizontal",
        legend.key.size = unit(1.5, "cm")) +
  labs(x = "Arrival Date", y = "Avg Bookings",
       title = "Average bookings: Resort Hotel vs. Market Segment",
       subtitle = "July 2015 - August 2017",
       color = "") +
  guides(color = guide_legend(nrow = 1)) +
  scale_color_manual(values = cbbPalette)
```

# Year-to-year Analysis

```{r}

```

