---
title: "ikea_eda"
author: "Sergio Puno"
date: "2/22/2022"
output: html_document
editor_options:
  chunk_output_type: console
---
Source:
https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-11-03/readme.md

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(tidytuesdayR)
library(scales)

```

# Load in the weekly data, IKEA information for this week.

Download weekly data, available in the `tt` object

```{r Load}
tt <- tt_load("2020-11-03")
```

```{r}
# checking for minor preprocessing steps required
glimpse(tt$ikea)

# first col isn't needed, just a row number/index
# additionally, pricing needs to be converted to USD, will add new column
# pricing based on date of 2020-11-03
ikea <- tt$ikea %>%
  select(-"...1") %>%
  mutate(price_usd = 0.27 * price,
         short_description = str_trim(str_replace_all(short_description,
                  "\\s+", " "))) %>%
  add_count(category, name = "category_total")

ikea %>%
  count(category, sort = TRUE) %>%
  mutate(category = fct_reorder(category, n)) %>%
  ggplot(aes(n, category)) +
  geom_col() +
  labs(x = "# of items",
       y = "",
       title = "Most common categories, IKEA items")

```

### Looking into the distribution of items and price: boxplot time.

Since the pricing can't go below 0, but can scale up to large values, we will
have to scale the x value of pricing (log 10). Pretty common to perform this
scaling on pricing information

```{r}
ikea %>%
  mutate(category = fct_reorder(category, price_usd)) %>%
  ggplot(aes(price_usd, category)) +
  geom_boxplot() +
  geom_jitter(width = 0, height = .1, alpha = 0.25) +
  scale_x_log10(labels = dollar) +
  labs(x = "Price (USD)",
       y = "",
       title = "Pricing per category")
```

This provides some insight into the variety of price offerings:
<ul>
<li> `Cafe furniture` is very niche in the product offerings in terms of the 
price range, especially compared to something like the `TV and media furniture`
category, which is more broad and has a lot of price points.
<li> With the ghost point included, you can get a sense of the number of total
products offered, `Cafe Furniture` again does not have nearly as many offerings 
as `Tables and desks` which shows a huge variety in products while still having
a fairly compact price range.
</ul>

We can look at this data in another way, showing the count of offerings with
the category name:

```{r}
library(glue)

ikea %>%
  mutate(category = glue("{ category } ({ category_total })"),
         fct_reorder(category, price_usd)) %>%
  ggplot(aes(price_usd, category)) +
  geom_boxplot() +
  scale_x_log10(labels = dollar) +
  labs(x = "Price (USD)",
       y = "",
       title = "Pricing per category")
```

Lets plot the same but using stacked density plots to check for bimodality, 
which would not be visible in a boxplot.

This would be possible with standard density plots with faceting but it might 
become a bit unruly so we will go with ggridges.
```{r}
library(ggridges)

ikea %>%
  mutate(category = glue("{ category } ({ category_total })"),
         fct_reorder(category, price_usd)) %>%
  ggplot(aes(price_usd, category)) +
  geom_density_ridges() +
  scale_x_log10(labels = dollar) +
  labs(x = "Price (USD)",
       y = "",
       title = "Pricing per category")
```
Let's get color options involved, the `other_colors` column is a True/False flag
so we can double up on the density plots and see if color has an impact on the
pricing distribution by category.

```{r}
ikea %>%
  mutate(category = glue("{ category } ({ category_total })"),
         fct_reorder(category, price_usd)) %>%
  ggplot(aes(price_usd, category, fill = other_colors)) +
  geom_density_ridges(alpha = 0.5) +
  scale_x_log10(labels = dollar) +
  labs(x = "Price (USD)",
       y = "",
       title = "Pricing per category")
```

An interesting take away here is the `Bar Furniture` category which shows in 
blue the availability of other colors as bimodal:
<ul>
<li> The higher price node has a similar median to the density in red for no 
other colors available. 
<li> The second node is distinctly lower price which may indicate that the 
availability of other colors may mean a change in material used or finish used
as many people prefer natural wood furniture which typically would not be
available in a variety of colors.
</ul>

The category with something similar but in the complete opposite is `Outdoor furniture`.
With this category the offerings with other colors as `True` have a higher number
available at a higher price.
### Product Names
Take a step back and let's look at the product naming to see if we need to 
do any cleaning

```{r}
ikea %>%
  count(name, category, sort = TRUE)
```
You can see, for example, the name `Besta` comes up for multiple categories, and
names come up many multiples of times within a category.

```{r}
ikea %>%
  mutate(name = fct_lump(name, 20)) %>%
  filter(name != "Other") %>%
  count(name, category, sort = TRUE) %>%
  mutate(name = fct_reorder(name, n, sum),
         category = fct_reorder(category, n, sum)) %>%
  ggplot(aes(n, name, fill = category)) +
  geom_col() +
  scale_fill_discrete(guide = guide_legend(reverse = TRUE)) +
  labs(x = "# of items",
       y = "Name of item")
```
This shows what a lot of people know about Ikea but in case you are unaware Ikea 
has names of product lines that typically will cover multiple categories.

For example, Besta might be the name of a storage system or storage brand which
offers tv stand units, wardrobe storage, cabinets, etc. 

## Short Description Analysis

```{r}
ikea %>%
  separate(short_description,
           c("main_description", "rest"),
           sep = ", ",
           extra = "merge",
           fill = "right",
           remove = FALSE) %>%
  extract(rest, "description_cm", "([\\d\\-xX]+) cm", remove = FALSE) %>%
  count(category, main_description, sort = TRUE)
```

Slight alteration to the above code

```{r}
ikea %>%
  separate(short_description,
           c("main_description", "rest"),
           sep = ", ",
           extra = "merge",
           fill = "right",
           remove = FALSE) %>%
  extract(rest, "description_cm", "([\\d\\-xX]+) cm", remove = FALSE) %>%
  unite(category_and_desc, category, main_description, sep = " - ") %>%
  count(category_and_desc, sort = TRUE)
```

### Lets look at Volume (size of furniture)

```{r}
ikea_volume <- ikea %>%
  select(item_id, name, short_description, category,
      price_usd, depth, height, width) %>%
  # 1 cubic cm is a milliliter, we can reprent the volume in liters
  # mutate(volume_liters = depth * height * width / 1000)
  # switching to cubic meters, divide by 1 million
  mutate(volume_m3 = depth * height * width / 1e6) %>%
  filter(!is.na(volume_m3),
          volume_m3 >= 0.001) %>%
  arrange(desc(volume_m3)) %>%
  add_count(category, name = "category_total")

# checking the volumes for low vol items to filter, retroactively adding filter
# to the above dataframe creation
ikea_volume %>%
  arrange(desc(volume_m3))

ikea_volume %>%
  mutate(category = glue("{ category } ({ category_total })"),
        category = fct_reorder(category, volume_m3)) %>%
  ggplot(aes(volume_m3, category)) +
  geom_boxplot() +
  scale_x_log10() +
  labs(x = "Volume of furniture (cubic meters, log scale)",
       y = "",
       title = "Volume of furniture by Category")
```

### Price vs. Volume
What is the highest price per cubic meter of furniture from Ikea?

* Impact of category vs impact of size

```{r}
ikea_volume %>%
  mutate(category = fct_lump(category, 6)) %>%
  ggplot(aes(volume_m3, price_usd, color = category)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10()
```

Calculate dollars per cubic meter

```{r}
ikea_volume %>%
  mutate(dollar_per_m3 = price_usd / volume_m3) %>%
  arrange(desc(dollar_per_m3))
```

### Looking at Designers

```{r}
ikea %>%
  group_by(designer) %>%
  summarise(n_items = n(),
            n_names = n_distinct(name),
            n_category = n_distinct(category)) %>%
  arrange(desc(n_names))
```

It would be difficult to build a model around designer as there are
many product names that cross the category and each will have a
relationship with pricing. For example, designer A might be focused
on lower cost models of tables, chairs, wardrobes, etc which would
impact the outcomes of the model.

```{r}
ikea_volume %>%
  lm(log2(price_usd) ~ log2(volume_m3), data = .) %>%
  summary()
```

Interpretation:



